---
title: "BIEN Data for mapping and biostatistics"
output: html_notebook
---

[BIEN](http://biendata.org)

Today, to build on our previous lesson with BIEN, we will re-visit some of the work we did to be sure proper libraries and some data are loaded. (As you know, no need to re-install.)

```{r}
#install.packages("BIEN")
library("BIEN")
library(ape) #Package for working with phylogenies in R
library(maps) #Useful for making quick maps of occurrences
library(sp) # A package for spatial data
#vignette("BIEN")
```

```{r}
# Download the data for a focal species.
Artocarpus_altilis <- BIEN_occurrence_species(species = "Artocarpus altilis")
```

It looks like nothing happened, but you actually have created a data frame called Artocarpus_altilis, full of character data and numeric data. Use the commands below. The first allows you to view the so called meta-data (data about the data; how many observations, for example). The second command looks at the first six entries in the data frame. The third looks at the last six entries in the data frame.

```{r}
str(Artocarpus_altilis)
head(Artocarpus_altilis)
tail(Artocarpus_altilis)
```

As you know from looking at the browser, you can look up a rangemap. 
```{r}
breadfruit_rangemap <- BIEN_ranges_load_species("Artocarpus altilis")
```

```{r}
# Map the range of breadfruit
map('world', fill=T, col="grey", bg = "light blue", xlim=c(-180,20),ylim=c(-60,80))
plot(breadfruit_rangemap, col="green", add=T)
```

```{r}
# Map the range of breadfruit + occurence records
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(breadfruit_rangemap, col="green", add=T)
points(cbind(Artocarpus_altilis$longitude, Artocarpus_altilis$latitude), col="blue", pch=20, cex=1)
```

By default, RBIEN commands for species occurrence only downloads limited information, but those defaults can be changed to get more expansive data, including:
- species in cultivation are added to those occurring naturally
- occurrence data outside of the Americas
- more taxonomy data (authority, higher levels than genus and species)
- when occurrence happened, country of occurrence, etc.

```{r}
Artocarpus_altilis_full <- BIEN_occurrence_species(species = "Artocarpus altilis",cultivated = T,only.new.world = F,all.taxonomy = T,native.status = T,observation.type = T,political.boundaries = T)

str(Artocarpus_altilis_full)
```

This was a slightly slower process because so much more data were obtained. 

Now let's map again, but on the world map instead of just the Americas. And let's show points from the full query in red, and overlay points from the default query (more limited) in blue. 

```{r}
map('world', fill=T , col="grey", bg="light blue")

#Plot the points from the full query in red
points(cbind(Artocarpus_altilis_full$longitude, Artocarpus_altilis_full$latitude), col="red", pch=20, cex=1)

# Plot the points from the default query in blue
points(cbind(Artocarpus_altilis$longitude, Artocarpus_altilis$latitude), col="blue", pch=20, cex=1) 
```

### Trait Data and linking this to Biostatistics

```{r}
BIEN_trait_list()
```

```{r}
# Create data frame of Persea traits
persea_traits <- BIEN_trait_genus(genus = "Persea")
str(persea_traits)
persea_traits
```

```{r}
Persea_DBH <- BIEN_trait_traitbygenus(trait = "diameter at breast height (1.3 m)", genus = "Persea")
head(Persea_DBH)
tail(Persea_DBH)
str(Persea_DBH)
```

```{r}
# List variable names from Persea_DBH data frame
names(Persea_DBH)
```

```{r}
Persea_DBH_small = Persea_DBH[c(1:5)]
Persea_DBH_small
Persea_DBH_small = as.data.frame(Persea_DBH_small)
class(Persea_DBH_small)
head(Persea_DBH_small)
class(Persea_DBH_small$trait_value)
DBH = as.numeric(Persea_DBH_small$trait_value)
class(DBH)
```

Note that a library is being loaded below, and without it the boxplots will not be generated. This should be a package installed by default when R is installed or updated, but if it's not then you will need to install it.

```{r}
library(lattice)
head(Persea_DBH_small)
boxplot(DBH~scrubbed_species_binomial, data=Persea_DBH_small)
```
  	
```{r}
Persea_WPH <- BIEN_trait_traitbygenus(trait = "whole plant height", genus = "Persea")
str(Persea_WPH)
Persea_WPH
```

```{r}
names(Persea_WPH)
```

```{r}
Persea_WPH_small = Persea_WPH[c(1:5)]
Persea_WPH_small
Persea_WPH_small = as.data.frame(Persea_WPH_small)
class(Persea_WPH_small)
head(Persea_WPH_small)
class(Persea_WPH_small$trait_value)
WPH = as.numeric(Persea_WPH_small$trait_value)
class(WPH)
```



```{r}
mean(DBH, data=Persea_DBH_small)
mean(WPH, data=Persea_WPH_small)
```


Downloading for just one species: 

```{r}
Persea_americana_DBH <- BIEN_trait_traitbyspecies(trait = "diameter at breast height (1.3 m)", species = "Persea americana")
head(Persea_americana_DBH)
tail(Persea_americana_DBH)
str(Persea_americana_DBH)
```


```{r}
Persea_americana_DBH_small = Persea_americana_DBH[c(1:5)]
Persea_americana_DBH_small
Persea_americana_DBH_small = as.data.frame(Persea_americana_DBH_small)
class(Persea_americana_DBH_small)
head(Persea_americana_DBH_small)
class(Persea_americana_DBH_small$trait_value)
DBHn = as.numeric(Persea_americana_DBH_small$trait_value)
class(DBHn)
```


```{r}
boxplot(DBHn)
hist(DBHn)
```
```{r}
boxplot(DBHn)
```

Another species: 


```{r}
Persea_lingue_DBH <- BIEN_trait_traitbyspecies(trait = "diameter at breast height (1.3 m)", species = "Persea lingue")
head(Persea_lingue_DBH)
tail(Persea_lingue_DBH)
str(Persea_lingue_DBH)
```


```{r}
Persea_lingue_DBH_small = Persea_lingue_DBH[c(1:5)]
Persea_lingue_DBH_small
Persea_lingue_DBH_small = as.data.frame(Persea_lingue_DBH_small)
class(Persea_lingue_DBH_small)
head(Persea_lingue_DBH_small)
DBHh = as.numeric(Persea_lingue_DBH_small$trait_value)
class(Persea_lingue_DBH_small$trait_value)
class(DBHh)
```



```{r}
library(dplyr)
Persea2<-rbind(Persea_americana_DBH_small, Persea_lingue_DBH_small)
head(Persea2)
tail(Persea2)
class(Persea2$trait_value)
DBHboth = as.numeric(Persea2$trait_value)
class(DBHboth)
```

```{r}
boxplot(DBHboth)
hist(DBHboth)
```

```{r}
PerseaBoth<-cbind(Persea2, DBHboth)
head(PerseaBoth)
tail(PerseaBoth)
```



```{r}
boxplot(DBHboth~scrubbed_species_binomial, data=PerseaBoth)
```
```{r}
DBHn
```
```{r}
DBHh
```
```{r}
ttest_result1 = t.test(x=DBHn, y=DBHh, paired=FALSE, alternative="two.sided")
ttest_result1
```
